% !TEX root = ../../../book.tex

In Part~I, our goal is to provide a review of some basic and some advanced statistical methodologies that are useful for developing trading algorithms. We begin with time series models (in Chapter~\ref{ch:ch_uvts}) for univariate data and provide a broad discussion of autoregressive, moving average models---from model identification, estimation, inference to model diagnostics. The stock price and return data exhibit some unique features and so we identify certain stylized facts regarding their behavior that have been empirically confirmed; this work will greatly help to discern any anomalies as and when they arise, as these anomalies generally indicate deviation from efficient market hypothesis. Although for modeling price and return data, only lower order models are needed, increasingly other trading features such as volume, volatility are being used in developing trading strategies. In particular,  predicting future volume flow to determine when to enter the market, requires the use of higher order models. Wherever possible we illustrate the methodology with examples and provide codes for computing, making the data accessible to the reader. 


This chapter is followed by methodologies for multiple time series data and the discussion in Chapter~\ref{ch:ch_mvts} is useful from pairs trading to portfolio optimization. The last chapter (Chapter~\ref{ch:ch_advanced}) in Part~I contains advanced topics such as theory of point processes as trading takes place on a continual basis. Understanding the market microstructure is essential for deciding about the entry to and exit from the market. This chapter also contains other modern topics such as machine learning and artificial intelligence. 


A reader with a strong statistics background can afford to skip some sections but we recommended to peruse these chapters as they contain discussions on topics that need further research work. In presenting the methodologies here, we keep the discussion lucid but immensely relevant to the main theme of this book---understanding the market behavior and developing effective trading strategies. 