% !TEX root = ../../../book.tex

In this part of the book we cover the other core topic of this book: Execution Strategies. Once again we will strive to give the reader the practitioner perspective and context, the necessary building blocks to fully grasp the subject and finally the quantitative underpinning where available.  Our hope is that by the end of the chapter our readers would have enough understanding of the topic to be able to, if asked, to implement at least a basic implementation of an execution product and be able to create the supporting Pre and Post-trade analytics.

We start in Chapter~\ref{ch:ch_trade_data_models} by discussing the modeling of trade data. We focus on several fundamental topics that are critical to practitioners but often disregarded by other treatment of the subject. We provide details on many of the analytics used in practice. Chapter ~\ref{ch:ch_mi_models} then goes into one of the fundamental subject of execution: Market Impact. We strive to bring in the intuition on the inner works of market impact and build the functional form of standard MI models from first principles. We then go into more details through a review of the sizable literature to guide the student that is interested in more depth. Finally in Chapter ~\ref{ch:ch_exec_models} goes into the real meat of the topic by covering Execution strategies and its component. Here we strive to give a modern perspective that tries to address many of the misconceptions present in the industry with the hope of setting the reader in more solid footing from the start.

Let's get started!
