% !TEX root = ../../book.tex
\pagestyle{empty}
%\pagenumbering{gobble}
%\clearpage


% Pre-back Page
\noindent {\itshape ``This work does a marvelous job of emphasizing the dual significance of determining the fair value of an asset as well as designing the optimal way to interact with the markets. Optimizing valuation is equally important to optimizing order execution. Both skills must be mastered to avoid selection bias and capturing value. This book must be read!''} \par
{%
\begin{flushright}
\begin{minipage}{0.6\textwidth}
\textbf{Peter J. Layton}, Principal, Blackthorne Capital Management, LLC (BCM)
\end{minipage}
\end{flushright} 
\twomedskip
}


\noindent {\itshape ``An outstanding and timely synthesis of the state of art algorithmic trading ideas. I will recommend it to all who is serious on the foundations.''} \par
{%
\begin{flushright}
\begin{minipage}{0.6\textwidth}
\textbf{Guofu Zhou}, Frederick Bierman and James E. Spears, Professor of Finance, Olin Business School (OBS), Washington University in St. Louis 
\end{minipage}
\end{flushright}
} \medskip



\newpage



% Back Page
\noindent\textbf{Algorithmic Trading and Quantitative Strategies} provides an in-depth overview of this growing field with a unique mix of quantitative rigor and practitioner's hands-on experience. The focus on both empirical modeling as well as practical know-how makes this book a valuable resource for students and professionals. \twomedskip


\noindent The book starts with the often overlooked context of why and how we trade via a detailed introduction to market structure and quantitative microstructure models. The authors then present the necessary quantitative toolbox including more advanced machine learning models needed to successfully operate in the field. They next discuss the subject of quantitative trading, alpha generation, active portfolio management and more recent topics like news and sentiment analytics. The last main topic of execution algorithms is covered in detail with emphasis on the state of the field including the elusive concept of market impact. The book concludes with a discussion on the technology infrastructure necessary to actually implement algorithmic strategies in large scale production settings. \twomedskip


\noindent A GitHub repository with explanatory and exercise notebooks and data-sets supplementing the book further helps bringing the material to live. \twomedskip


\noindent Raja Velu is a Professor of Finance and Business Analytics in the Whitman School of Management at Syracuse University. He served as a Technical Architect at Yahoo! in the Sponsored Search Division and was a visiting scientist at IBM-Almaden, Microsoft Research, Google and JPMC. He has also held visiting positions at Stanford, Indian School of Business, National University of Singapore and Singapore Management University. \medskip

\noindent Maxence Hardy is an Executive Director and the Head of eTrading Quantitative Research for Equities and Futures at J.P.Morgan, based in New York. Mr. Hardy is responsible for the development of the algorithmic trading strategies for the Equities and Futures divisions globally. \medskip

\noindent Daniel Nehren is a Managing Director and the Head of Statistical Modelling and Development for Equities at Barclays. Based in New York, Mr. Nehren is responsible for the development of algorithmic trading and analytics products. Mr. Nehren has over 19 years of experience in equity trading working for some of the most prestigious financial firms including Citadel, J.P Morgan, and Goldman Sachs. 