% !TEX root = ../../exercises.tex
\chapter{Advanced Topics \label{ch:ch_advanced}}


The dataset consists of (CSV) file that contains the entire trading session, including early and late hours from 6:00~a.m. to 8:00~p.m. EST. It contains intraday depth book activity several major tickers from up to 5 major national exchanges: Nasdaq, Direct Edge, NYSE, ARCA, and BATS (exact number of exchanges depends on the historical period). All messages are consolidates into one file ordered by timestamp. This dataset allows to build a full national depth book (super book) at any moment intraday. 

\begin{table}[h!]
   \caption{Column Format}
   \centering
   \begin{tabular}{p{2cm}p{8cm}} 
   \textbf{Column} & \text{Description} \\ \hline
   Timestamp & Number of milliseconds after the midnight. \\ \hline
   Ticker & Equity symbol (up to 8 characters) \\ \hline
   Order & Unique order ID. \\ \hline
   T & Message type. Allowed Values: \newline
	~~\llap{\textbullet}~~ ``B''--Add buy order \newline
	~~\llap{\textbullet}~~ ``S''--Add sell order \newline
	~~\llap{\textbullet}~~ ``E''--Execute outstanding order in part \newline
	~~\llap{\textbullet}~~ ``F''--Execute outstanding order in full \newline
	~~\llap{\textbullet}~~ ``D''--Delete outstanding order in full \newline 
	~~\llap{\textbullet}~~ ``X''--Bulk volume for the cross event \newline
	~~\llap{\textbullet}~~ ``T''--Execute non-displayed order  \\ \hline
   Shares & Order quantity, available for the ``B'', ``S'', ``E'', ``X'', ``C'', ``T'' messages. Zero for ``F'' and ``D'' messages. \\ \hline
   Price & Order price, available for the ``B'', ``S'', ``X'', and ``T'' order messages. Zero for cancellation and executions. The last 4 digits are decimal digits. The decimal portion is padded on the right with zeros. The decimal point is implied by position; it does not appear inside the price field. Divide by 10000 to convert into currency value. \\ \hline
   MPID & Market Participant ID associated with the transaction (4 characters) \\ \hline
   MCID &U Market Center Code (originating exchange--1 character)
   \end{tabular}
\end{table}


For the analysis of Problems 1--3, we consider the data for CISCO. \\


% Problem
\prob You need to aggregate the data into 5-min intervals before answering the following.
        \begin{enumerate}[(a)]
        \item Let $x_t$ denote the number of trades in the $t$th 5-minute interval. Ignore the time gaps between trading days. Plot the time series and its ACF. Determine if there are intraday period patterns in the series.
        \item Using the last transaction in the $t$th 5-minute interval as the stock price in that interval, plot the time series $y_t$ of 5-minute returns during the period and the corresponding ACF.
        \item Consider the bivariate time series ($x_t$, $y_t$). How does $y_t$ vary with $x_t$? Are there intraday periodic patterns in ($x_t$, $y_t$)?
        \item Plot the durations of the transaction times for Mondays and Fridays. Is there any difference in the patterns?
        \item Fit GARCH models for the durations and interpret the coefficients.
        \item Is there any particular exchange that offers better price? \\
        \end{enumerate}


% Problem
\prob Use the transaction data in Problem~1. There are seventy-eight five-minute intervals in a trading day. Let $d_i$ be the average of all log durations for the $i$th 5-minute interval across all trading days. Let $t_j$ be tick time of a trade during the $i$th 5-minute interval, define an adjusted duration as $\Delta t_j^* = \Delta t_j/\exp(d_i)$ where $\Delta t_j = t_j - t_{j-1}$.
        \begin{enumerate}[(a)]
        \item Is there a diurnal pattern in the adjusted duration series?
        \item Build EACD and WACD models for the adjusted duration and compare them. Now aggregate the data over the 5-minute intervals; Let $r_t$ be the return and $V_t$ is the volume of transactions in the $t$th interval.
        \item Is there any relationship between the volume $V_t$ and volatility $r_t^2$ . If there is develop a model that will capture the dependence of volatility on volume. Let $\overline{p}_t$ be the average price in the $t$th 5-minute
        interval. \\
        \end{enumerate}


% Problem
\prob Use the CISCO execution data only; aggregate at various time intervals: 1~min, 2~min, \dots, 15~min. For these levels record
        \begin{enumerate}[(a)]
        \item Average price, VWAP, realized volatility and aggregate volume.
        \item By relating the returns and volatility estimates based on average price to realized volatility, identify the optimal duration for aggregation that will cancel out the market friction but pick up the true information. (You may want to refer to Bandi and Russell (2006)~\cite{bandi}). \\
        \end{enumerate}


\addtocounter{total}{\theproblem}